Keep your Laziness in Check

### 2018-03-09 paper/main.tex
[paper] minor edits

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
The \lk{Spec} constructor takes a function, whose first
argument \lk{predict} is the predicted demand on the \lk{Int} argument
of \lk{take}, and second argument the predicted demand on the \lk{[a]}
argument. This spec simply says that the \lk{Int} argument will always
be used, and the demand on the list argument is either the same as the
demand on the output, or demand on the output plus some
unevaluated \lk{thunk} at the end. The former case happens when $n$ is
larger than the list's length, otherwise the latter case applies.

@@

Amongst his work on purely functional data structures\cite{Okasaki},
Okasaki cleverly improved upon this design using lazy lists that both
incrementally performs the reversal of the back list, and
incrementally appends the reversal of the back list onto the front
list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
  rotate :: [a] -> [a] -> [a] -> [a]
  rotate      []       []  as =                       as
  rotate      []  (b : bs) as =     rotate [] bs (b : as)
  rotate (f : fs)      []  as = f : rotate fs []      as
  rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, unlike
the naive reverse-and-append operation, \lk{rotate} is much
lazier. A \lk{reverse} function has to evaluate the entire structure
of the back list, which is not incremental. On the other
hand, \lk{rotate} produces a list immediately, and packs the rest of
the computation into the lazy list \lk{rest}.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
this expression, the first two underscores represent unevaluated
elements in the list, while the last underscore represents unevaluated
tail of the list.

@@

Putting these pieces together, we can specify the laziness behavior
of \lk{rotate} with Okasaki's insight: as long as the front list is
longer than the back list, demanding the first element from the result
of \lk{rotate} only causes evaluation of the first element of the
front list ~\cite{Okasaki95simpleand}. Since \lk{rotate} is always
used by supplying the empty list \lk{[]} as the accumulator argument,
we define \lk{rot fs bs = rotate fs bs []}, and show a specification
of \lk{rot}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rotate\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}

@@

Moreover, it is important to be able to distinguish whether such a datatype
follows a desired laziness behavior. For instance, the standard functional
solution for dynamic programming problems crucially uses laziness \hzh{is this
solution really standard? Is there some work we can cite to establish this
approach as the canonical approach for dynamic programming in functional
programming?}. Let us consider the famous knapsack problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

@@

To catch this behavior through testing, we can write another specification that
captures this correctness condition. The well-foundedness \hzh{is this a word?}
of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
of the map is demanded. Let $\lk{solutionStep}_n$ be the $n$-times iterated
step function \lk{solutionStep . solutionStep . ... . solutionStep}, then if we
demand the value at \lk{(j, k)} from the result of $\lk{solutionStep}_n$, its
input map should not be evaluated at the \lk{(j, k)} entry since otherwise
evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.

@@

iterSolution_spec :: Evaluation '[Key, Int, KMap] (Maybe Int)
                  -> Maybe (Int, Int)
iterSolution_spec (Evaluation args demands dOut) =
  let I (Key evalK) = hd args
      I nIter       = hd (tl args)
      dInM          = hd (tl (tl demands))
      inM           = replaceThunk Empty (fromDemand @KMap dInM)
  in  if (inM == Empty)   ||
         isBaseCase evalK ||
         nIter <= 0       ||
         (isNothing $ lookup inM evalK)
      then Nothing
      else Just evalK
  where isBaseCase (j, k) = j - 1 < 0 || k - weights !! j < 0
\end{inlinecode}
%
This specification is different from \lk{rot\_spec} --- it uses
the \lk{Evaluation} type instead of \lk{Spec}. In StrictCheck, we
reverse the \lk{Spec} type for specifications that computes the exact
demand on inputs. However, there many cases where we might want to
only check whether some property holds of the observed demands on the
inputs. For example, here we only need to check that the value
at \lk{(j, k)} is not evaluated, and we don't compute the exact demand
on the input map. For these specifications, we write a function that
takes an \lk{Evaluation} value, which contains the inputs, the
observed demand on the inputs, and a randomly generated demand on the
output by the framework.

>>>>

How, then, can we test and catch such laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following example using the standard
\lk{take} function on lists, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict n (if n > length xs then d else d ++ thunk)
\end{inlinecode}
%
The \lk{Spec} constructor takes a function, whose first
argument \lk{predict} is is a continuation that takes the predicted
demands on the inputs to \lk{take}, and second argument of the
function is the observed demand on the output \lk{[a]}. This spec says
that the \lk{Int} argument will always be used, and the demand on the
list argument is either the same as the demand on the output, or
demand on the output plus some unevaluated \lk{thunk} at the end. The
former case happens when $n$ is larger than the list's length,
otherwise the latter case applies.

@@

Amongst his work on purely functional data structures\cite{Okasaki},
Okasaki cleverly improved upon this design using lazy lists that both
incrementally performs the reversal of the back list, and
incrementally appends the reversal of the back list onto the front
list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
  rotate :: [a] -> [a] -> [a] -> [a]
  rotate      []       []  as =                       as
  rotate      []  (b : bs) as =     rotate [] bs (b : as)
  rotate (f : fs)      []  as = f : rotate fs []      as
  rotate (f : fs) (b : bs) as = f : rotate fs bs (b : as)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, unlike
the naive reverse-and-append operation, \lk{rotate} is much
lazier. A \lk{reverse} function has to evaluate the entire structure
of the back list, which is not incremental. On the other
hand, \lk{rotate} produces a list immediately, and packs the rest of
the computation into the tail of the lazy list.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the evaluated part of a
data structure. For lists, \textit{demands} will be prefixes of
lists. We use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} instead of explicit demand type values. In
this expression, the first two underscores represent unevaluated
elements in the list, while the last underscore represents unevaluated
tail of the list.

@@

Putting these pieces together, we can specify the laziness behavior
of \lk{rotate} with Okasaki's insight: as long as the front list is
longer than the back list, demanding the first element from the result
of \lk{rotate} only causes $O(1)$ amount of evaluation
~\cite{Okasaki95simpleand}. Since \lk{rotate} is always used by
supplying the empty list \lk{[]} as the accumulator argument, we
define \lk{rot fs bs = rotate fs bs []}, and show a specification
of \lk{rot}.
%
\begin{inlinecode}
rot_spec :: Shaped a => Spec '[[a], [a]] [a]
rot_spec =
  Spec $ \predict d fs bs ->
    let (fs', bs') = splitAt (length fs) (cap d)
        spineLen  = length (cap (d ++ [undefined]))  -- # of spine thunks forced
        overflow  = spineLen       > length fs  -- begun taking from bs?
        overrot   = length (cap d) > length bs  -- forced all of bs?
        padLength =
          length bs `min`
            if overflow
            then length bs - length bs'
            else length (cap d)
        spinePad = replicate padLength thunk
    in predict
         (                    fs' ++ if overflow            then [] else thunk)
         (spinePad ++ reverse bs' ++ if overflow || overrot then [] else thunk)
\end{inlinecode}
%
The \lk{rot\_spec} exactly captures the core idea from Okasaki's \lk{rotate}
operation, and testing the reverse and append implementation against this
specification immediately produces a counter example.
\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{revappend_counterexample}
\caption{Counterexample for \lk{revAppend} under \lk{rot\_spec}}
\end{figure}

@@

Moreover, it is important to be able to distinguish whether such a
datatype follows a desired laziness behavior. \hzh{the previous
sentence and the next one feels disconnected} For instance, the
standard functional solution for dynamic programming problems
crucially uses laziness \hzh{is this solution really standard? Is
there some work we can cite to establish this approach as the
canonical approach for dynamic programming in functional
programming?}. Let us consider the famous knapsack problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

@@

To catch this behavior through testing, we can write another
specification that captures this correctness condition. Termination
of \lk{solution} depends on what is evaluated when the value at
key \lk{(j, k)} of the map is demanded. Let $\lk{solutionStep}_n$ be
the $n$-times iterated step function \lk{solutionStep . solutionStep
. ... . solutionStep}, then if we demand the value at \lk{(j, k)} from
the result of $\lk{solutionStep}_n$, its input map should not be
evaluated at the \lk{(j, k)} entry since otherwise evaluating the
value at \lk{(j, k)} becomes an infinite loop at the fixpoint.

@@

iterSolution_spec :: Evaluation '[Key, Int, KMap] (Maybe Int)
                  -> Maybe (Int, Int)
iterSolution_spec (Evaluation args demands dOut) =
  let I (Key evalK) = hd args
      I nIter       = hd (tl args)
      dInM          = hd (tl (tl demands))
      inM           = replaceThunk Empty (fromDemand @KMap dInM)
  in  if (inM == Empty)   ||
         isBaseCase evalK ||
         nIter <= 0       ||
         (isNothing $ lookup inM evalK)
      then Nothing
      else Just evalK
  where isBaseCase (j, k) = j - 1 < 0 || k - weights !! j < 0
\end{inlinecode}
%
This specification is different from \lk{rot\_spec} --- it uses
the \lk{Evaluation} type instead of \lk{Spec}. In StrictCheck, we
reserve the \lk{Spec} type for specifications that computes the exact
demand on inputs. However, there many cases where we might want to
only check whether some property holds of the observed demands on the
inputs. For example, here we only need to check that the value
at \lk{(j, k)} is not evaluated, and we don't compute the exact demand
on the input map. For these specifications, we write a function that
takes an \lk{Evaluation} value, which contains all participants
necessary for analyzing laziness properties of \lk{iterSolution}.
